Clustering Constraints on the Relative Sizes of Central and Satellite Galaxies

### 2017-11-20 paper/galsize_paper.tex
Updated draft with revisions suggested by Ben Moster

\author[Hearin, Behroozi, Kravtsov \& Moster]{
Andrew Hearin$^{1}$, Peter Behroozi$^{2}$, Andrey Kravtsov$^{3}$, Benjamin Moster$^{4}$\\
$^{1}$Argonne National Laboratory, Argonne, IL, USA 60439, USA\\
$^{2}$Department of Physics, University of Arizona, 1118 E 4th St, Tucson, AZ 85721 USA\\
$^{3}$Department of Astronomy \& Astrophysics, and Kavli Institute for Cosmological Physics, University of Chicago, Chicago IL 60637\\
$^{4}$Universit{\"a}ts-Sternwarte, Ludwig-Maximilians-Universit{\"a}t M{\"u}nchen, Scheinerstr. 1, 81679 M{\"u}nchen, Germany
}

@@

As the bedrock of our modeling, we use the publicly available\footnote{\url{http://www.slac.stanford.edu/~behroozi/BPlanck\_Hlists}} catalog of {\tt Rockstar} subhalos identified at $z=0$ in the Bolshoi-Planck simulation \citep{klypin_etal11,klypin_etal16,behroozi12_rockstar,riebe_etal13,behroozi_etal12b,rodriguez_puebla16_bolplanck}. As described in \S\ref{subsec:sham}, we use traditional abundance matching to connect stellar mass $\mstar$ with subhalo mass. To address issues related to subhalo incompleteness \citep{guo_white13,campbell_etal17}, we supplement the Bolshoi-Planck subhalo catalog with subhalos that have been disrupted and no longer appear in the standard catalog as $z=0$ surviving subhalos. We describe our treatment of these ``orphan" subhalos in the Appendix, where we also show results that exclude all orphan subhalos.

For our SDSS galaxy samples, we calculate two-point clustering $\wproj$ using line-of-sight projection of $\pi_{\rm max}=20~\mpc$ using the {\tt correl} program in {\tt UniverseMachine}. The {\tt correl} code uses the \citet{landy_szalay93} estimator ($DD - 2DR + RR$) to compute the redshift-space correlation function $\xi(\rproj, \pi)$, which is then integrated over $\vert\pi\vert < 20$ Mpc to compute the projected correlation function $\wproj(\rproj)$.  The code uses $10^6$ randoms drawn from the same mask region with uniform volume distribution to compute $DR$, and $RR$ is computed via Monte Carlo integration. Errors are estimated by jackknife resampling, giving us an estimate for the lower bound of samples with volumes $V_{\rm eff}\lesssim 0.3{\rm Gpc}^3.$

@@

We map $\mstar$ onto subhalos using deconvolution abundance matching (SHAM). See Section 3.3 of \citet{behroozi_etal10}, or the appendix of \citet{kravtsov_etal14}, for an account of the mathematics underlying this technique in the presence of scatter. To perform the abundance matching, we use the publicly available code\footnote{\url{https://bitbucket.org/yymao/abundancematching/overview}} developed by Yao-Yuan Mao \citep{lehmann_etal15} that provides a python wrapper around the C kernel developed in \citet{behroozi_etal10}. For the subhalo property used in SHAM, we use $\mpeak,$ the largest value of $\mvir$ ever attained by the subhalo throughout the entire history of the main progenitor halo.  Using the stellar mass function provided in \citet{moustakas_etal13}, we model $\mstar$ as a log-normal distribution with $0.2$dex of scatter about the median relation $\median{\mstar}{\mpeak}$ determined by SHAM \citep{reddick_etal13}. As described in the appendix, we include a prescription for supplementing the ordinary {\tt Rockstar} subhalo catalog with disrupted subhalos, so that some of our model galaxies inhabit subhalos that are no longer resolved and do not appear in the standard publicly available catalog.

Figure \ref{fig:baseline_sham_clustering} shows that our implementation of SHAM produces model galaxies whose projected clustering is in reasonably good agreement with SDSS galaxies. We note, however, that $\wproj(\rproj\gtrsim1{\rm Mpc})$ is $\sim10-20\%$ too weak at stellar mass $\mstar\gtrsim10^{10.75}\msun,$ and similarly on all scales for the $\mstar\gtrsim10^{9.75}\msun$ sample, although the statistical significance of this discrepancy is likely rather mild due to the sample variance of SDSS galaxies with $z\lesssim0.05.$

@@

In Figure \ref{fig:scatter_plot} we show the scaling of galaxy size $\rhalf$ with $\mstar.$ The black curve enveloped by the gray bands shows the scaling relation for SDSS galaxies, while the blue curve shows the median relation  $\median{\rhalf}{\mstar}$ predicted using the $\rvir-$based model described in \S\ref{subsubsec:rvirmodel}. This figure shows that models in which $\rhalf\propto\rvir$ can naturally match the observed $\rhalf-\mstar$ relation, confirming the results of \citet{kravtsov13} in a forward modeling context.

@@

Our methodology is closely aligned with \citet{somerville_etal17}, who studied the empirical modeling features that are necessary to recover the tight scatter in the observed $\mean{\rhalf}{\mstar}$ relation. By building models where $\rhalf$ is set by halo spin $\lambda_{\rm halo}$, it was found in \citet{somerville_etal17} that the level of intrinsic scatter about $\langle\lambda_{\rm halo}\vert\mhalo\rangle$ in dark matter halos is at least as large as the scatter about $\langle\rhalf\vert\mstar\rangle$ seen in observed galaxies. In our approach, the level of scatter is simply a modeling parameter whose fiducial value was motivated by \citet{somerville_etal17}. In ongoing follow-up work discussed in \S\ref{sec:future}, we will systematically test the large-scale structure implications of the assumption that $\lambda_{\rm halo}\propto\rhalf^{\rm disk}$ using forward-modeling methods analogous to those advocated for in \citet{somerville_etal17} and employed here.

@@

On the one hand, this degeneracy with the satellite fraction is unfortunate, because it means galaxy size does not leave a pure and unique signature on $\rhalf-$clustering ratios. However, this can also be viewed as an opportunity to extract tighter constraints on $F_{\rm sat}(\mstar),$ which are sorely needed to discriminate between competing models \citep{watson_conroy13}. Traditional galaxy clustering is already being used to validate and/or fit models of the stellar-to-halo-mass relation \citep[e.g.,][]{leauthaud_etal11,moster_etal13,behroozi13_smhm,lehmann_etal15}. Based on our results, we advocate that empirical models for $\mean{\mstar}{\mhalo}$ be supplemented with additional model ingredients for $\median{\rhalf}{\mhalo},$ and that the parameters of the composite model be {\em jointly} constrained by measurements of the stellar mass function, $\mstar-$dependent clustering, and $\rhalf-$dependent clustering. Because the number of free parameters required to additionally model galaxy size are so few, such analyses will be able to exploit the additional statistical leverage provided by the $\rhalf-$dependence of $\wproj(\rproj)$ to significantly improve constraints on $F_{\rm sat}(\mstar),$ without any appeal to a group-finder.

@@

APH thanks John Baker for the {\em Toejam \& Earl} soundtrack. Thanks also to Frank van den Bosch, Andrew Zentner, Doug Watson, and Risa Wechsler for thoughtful feedback at various stages of the development of this work, to Faustin Carter and Sebastian Bocquet for sharing their matplotlib expertise, and to Andrew Zentner for Scottish grammar consultation.

@@

Work done at Argonne National Laboratory was supported under the DOE contract DE-AC02-06CH11357. This research was supported in part by the National Science Foundation under Grant No. NSF PHY11-25915.  AK was  supported by NSF grant  AST-1412107, and by the Kavli Institute for Cosmological Physics at the University of Chicago through grant PHY-1125897, and an endowment from the Kavli Foundation and its founder Fred Kavli.

>>>>

\author[Hearin, Behroozi, Kravtsov \& Moster]{
Andrew Hearin$^{1}$, Peter Behroozi$^{2}$, Andrey Kravtsov$^{3}$, Benjamin Moster$^{4,5}$\\
$^{1}$Argonne National Laboratory, Argonne, IL, USA 60439, USA\\
$^{2}$Department of Physics, University of Arizona, 1118 E 4th St, Tucson, AZ 85721 USA\\
$^{3}$Department of Astronomy \& Astrophysics, and Kavli Institute for Cosmological Physics, University of Chicago, Chicago IL 60637\\
$^{4}$Universit{\"a}ts-Sternwarte, Ludwig-Maximilians-Universit{\"a}t M{\"u}nchen, Scheinerstr. 1, 81679 M{\"u}nchen, Germany\\
$^5$Max-Planck Institut f\"ur Astrophysik, Karl-Schwarzschild Stra\ss e 1, 85748 Garching, Germany
}

@@

As the bedrock of our modeling, we use the publicly available\footnote{\url{http://www.slac.stanford.edu/~behroozi/BPlanck\_Hlists}} catalog of {\tt Rockstar} subhalos identified at $z=0$ in the Bolshoi-Planck simulation \citep{klypin_etal11,klypin_etal16,behroozi12_rockstar,riebe_etal13,behroozi_etal12b,rodriguez_puebla16_bolplanck}. As described in \S\ref{subsec:sham}, we use traditional abundance matching to connect stellar mass $\mstar$ with subhalo mass. To address issues related to subhalo incompleteness \citep{guo_white13,campbell_etal17}, we supplement the Bolshoi-Planck subhalo catalog with subhalos that have been disrupted and no longer appear in the standard catalog as $z=0$ surviving subhalos. We describe our treatment of these ``orphan" galaxies in the Appendix, where we also show results that exclude all orphan galaxies.

For our SDSS galaxy samples, we calculate two-point clustering $\wproj$ using line-of-sight projection of $\pi_{\rm max}=20~\mpc$ using the {\tt correl} program in {\tt UniverseMachine} (Behroozi et al. 2017, in prep). The {\tt correl} code uses the \citet{landy_szalay93} estimator ($DD - 2DR + RR$) to compute the redshift-space correlation function $\xi(\rproj, \pi)$, which is then integrated over $\vert\pi\vert < 20$ Mpc to compute the projected correlation function $\wproj(\rproj)$.  The code uses $10^6$ randoms drawn from the same mask region with uniform volume distribution to compute $DR$, and $RR$ is computed via Monte Carlo integration. Errors are estimated by jackknife resampling, giving us an estimate for the lower bound of samples with volumes $V_{\rm eff}\lesssim 0.3{\rm Gpc}^3.$

@@

We map $\mstar$ onto subhalos using deconvolution abundance matching (SHAM). See Section 3.3 of \citet{behroozi_etal10}, or the appendix of \citet{kravtsov_etal14}, for an account of the mathematics underlying this technique in the presence of scatter. To perform the abundance matching, we use the publicly available code\footnote{\url{https://bitbucket.org/yymao/abundancematching/overview}} developed by Yao-Yuan Mao \citep{lehmann_etal15} that provides a python wrapper around the C kernel developed in \citet{behroozi_etal10}. For the subhalo property used in SHAM, we use $\mpeak,$ the largest value of $\mvir$ ever attained by the subhalo throughout the entire history of the main progenitor halo.  Using the stellar mass function provided in \citet{moustakas_etal13}, we model $\mstar$ as a log-normal distribution with $0.2$dex of scatter about the median relation $\median{\mstar}{\mpeak}$ determined by SHAM \citep{reddick_etal13}. As described in the appendix, we include a prescription for supplementing the ordinary {\tt Rockstar} subhalo catalog with disrupted subhalos, so that some of our model galaxies inhabit subhalos that are no longer resolved and do not appear in the standard publicly available catalog (``orphan galaxies").

Figure \ref{fig:baseline_sham_clustering} shows that our implementation of SHAM produces model galaxies whose projected clustering is in reasonably good agreement with SDSS galaxies. We note, however, that small-scale clustering shows mild $\sim10-20\%$ tension at stellar mass $\mstar\gtrsim10^{10.75}\msun,$ and similarly on all scales for the $\mstar\gtrsim10^{9.75}\msun$ sample, although the statistical significance of this discrepancy is likely rather mild due to the sample variance of SDSS galaxies with $z\lesssim0.05.$

@@

In Figure \ref{fig:scatter_plot} we show the scaling of galaxy size $\rhalf$ with $\mstar.$ The black curve enveloped by the gray bands shows the scaling relation for SDSS galaxies, while the blue curve shows the median relation  $\median{\rhalf}{\mstar}$ predicted using the $\rvir-$based model described in \S\ref{subsubsec:rvirmodel}. While the curvature of the median relation in the model is slightly weaker than SDSS, this figure shows that models in which $\rhalf\propto\rvir$ can nonetheless match the observed $\rhalf-\mstar$ relation remarkably well given only a single parameter, confirming the results of \citet{kravtsov13} in a forward modeling context.

@@

Our methodology is closely aligned with \citet{somerville_etal17}, who studied the empirical modeling features that are necessary to recover the tight scatter in the observed $\mean{\rhalf}{\mstar}$ relation. By building models where $\rhalf$ is set by halo spin $\lambda_{\rm halo}$, it was found in \citet{somerville_etal17} that the level of intrinsic scatter about $\langle\lambda_{\rm halo}\vert\mhalo\rangle$ in dark matter halos is at least as large as the scatter about $\langle\rhalf\vert\mstar\rangle$ seen in observed galaxies. In our approach, the level of scatter is simply a modeling parameter whose fiducial value was motivated by \citet{somerville_etal17}. In ongoing follow-up work discussed in \S\ref{sec:future}, we will systematically test the large-scale structure implications of the assumption that $\rhalf^{\rm disk}\propto\lambda_{\rm halo}$ using forward-modeling methods analogous to those advocated for in \citet{somerville_etal17} and employed here.

@@

On the one hand, this degeneracy with the satellite fraction is unfortunate, because it means galaxy size does not leave a pure and unique signature on $\rhalf-$clustering ratios. However, this can also be viewed as an opportunity to extract tighter constraints on $F_{\rm sat}(\mstar),$ which are sorely needed to discriminate between competing models \citep{watson_conroy13}. Traditional galaxy clustering is already being used to validate and/or fit models of the stellar-to-halo-mass relation \citep[e.g.,][]{leauthaud_etal11,moster_etal10,behroozi13_smhm,lehmann_etal15}. Based on our results, we advocate that empirical models for $\mean{\mstar}{\mhalo}$ be supplemented with additional model ingredients for $\median{\rhalf}{\mhalo},$ and that the parameters of the composite model be {\em jointly} constrained by measurements of the stellar mass function, $\mstar-$dependent clustering, and $\rhalf-$dependent clustering. Because the number of free parameters required to additionally model galaxy size are so few, such analyses will be able to exploit the additional statistical leverage provided by the $\rhalf-$dependence of $\wproj(\rproj)$ to significantly improve constraints on $F_{\rm sat}(\mstar),$ without any appeal to a group-finder.

@@

APH thanks John Baker for the {\em Toejam \& Earl} soundtrack. Thanks also to Frank van den Bosch, Andrew Zentner, Doug Watson, Simon White, Thorsten Naab, Risa Wechsler and Chihway Chang for thoughtful feedback at various stages of the development of this work, to Faustin Carter and Sebastian Bocquet for sharing their matplotlib expertise, and to Andrew Zentner for Scottish grammar consultation.

@@

Work done at Argonne National Laboratory was supported under the DOE contract DE-AC02-06CH11357. This research was supported in part by the National Science Foundation under Grant No. NSF PHY11-25915.  AK was  supported by NSF grant  AST-1412107, and by the Kavli Institute for Cosmological Physics at the University of Chicago through grant PHY-1125897, and an endowment from the Kavli Foundation and its founder Fred Kavli. BPM acknowledges an Emmy Noether grant funded by the Deutsche Forschungsgemeinschaft (DFG, German Research Foundation) -- MO 2979/1-1.
